Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-23 manuscript/manuscript.Rmd
finished addressing all minor comments by R2.

Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).

@@

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis[-@hutchence2021]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather inconsipicous. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue. It is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice[@milks2019].

>>>>

Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences. 2) How does training affect novices' performance in these two aspects? Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template.

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$). These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (skill level) and PC2 (mental template). However, the effects of training across different assessment periods on both dimensions are not significant.

@@

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue. Corroborating Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
